Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-09-24 paper/synthetic.tex
wip paper

We compare the performance of B2B against five competing methods:
%
Ridge regression \citep{hoerl1959optimum}, multitask Lasso \citep{argyriou2008convex}, Partial Least Squares or PLS \citep{wold_pls, tenenhaus_pls}, Canonical Correlation Analysis or CCA \citep{cca_hotelling}, and Reduced Rank Regression or RRR \citep{Izenman_rrr}.
%
Each of these methods estimates a matrix of coefficients $\hat{W} \in \mathbb{R}^{d_x \times d_y}$, from which we estimate the diagonal mask $\hat{E}$ using the feature importances $\hat{E}_{i,i} = \| W_{i, :} \|$.
%
For B2B, we obtain $\hat{E}$ as described in Algorithm~\ref{algorithm:b2br}.
%
We employ $5$-fold cross-validation to select the optimal hyper-parameters of each method.

@@

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\textwidth]{synthetic.pdf}
    \caption{Results of synthetic experiments. Each dot depicts the average value of a metric for B2B ($x$-axis) against a competitor ($y$-axis) for each experiment configuration averaged over $20$ random seeds. The prediction error on $Y$ within (in) and outside (out) of the training distribution are weakly but consistently larger for all of the models as compared to B2B with and without the Sonquist binarization (E). The AUC, measuring the ability of each model to separate causal from noncausal factors, is consistently larger in B2B as compared to the other models.}
    \label{table:synthetic}
\end{figure}

>>>>

We compare the performance of B2B against four competing methods, all implemented in scikit-learn \citep{scikit}:
%
% To be updated

\subsection{Baseline models}

Forward regression consists of an $l2$-regularized "ridge" regression from the putative causes $X$ to the observations $Y$:
\begin{equation}
  H = (X^T X +\lambda I)^{-1} X^T Y
\end{equation}

Backward regression consists of an $l2$-regularized "ridge" regression from $Y$ to $X$:
\begin{equation}
  G = (Y^T Y +\lambda I)^{-1} Y^T X
\end{equation}

CCA finds $G\in\mathbb{R}^{d_z, d_y}$ and $H\in\mathbb{R}^{d_z, d_x}$ such that $X$ and $Y$ are maximally correlated in a latent $Z$ space:
\begin{equation}
  max corr(XH^T, YG^T)
\end{equation}

% To be checked
PLS finds $G\in\mathbb{R}^{d_z, d_y}$ and $H\in\mathbb{R}^{d_z, d_x}$ such that $X$ and $Y$ are maximally covarying in a latent $Z$ space:
\begin{equation}
  max cov(XH^T, YG^T)
\end{equation}

We employ $5$-fold cross-validation to select the optimal number of components for CCA and PLS. Regressions were l2-regularized with a $\lambda$ regularization parameters fitted with the efficient leave-one-out procedure implemented in scikit-learn RidgeCV \citep{scikit}.

\subsection{Evaluating Causal Discovery from models' coefficients}

B2B leads to unbiased (i.e. zeros-centered) scalar coefficients for non-causal features. In contrast, the Forward, Backward, CCA and PLS models lead to a loading vector $H_i$ per feature $i$ (or one vector $G^i$ for the backward model). To transform such vector into an estimated causal contribution $\hat E$, we take the sum of square coefficients:

\begin{equation}
  \hat E_i = \sum_j {H^j_i}^2
\end{equation}

To estimate whether models accurately identify causal factors, we compute the area-under-the-curve (AUC) across factors $AUC(E, \hat E)$.
%\begin{equation}
%  AUC = 1 - \sum_1^n (E_k - E_{k-1}) ( \hat{E}_k + \hat{E}_{k-1}) / 2
%\end{equation}

\subsection{Evaluating Causal Discovery with held-out prediction reliability}

In most cases, $E$ is not known and AUC can thus not be estimated.

To address this issue, we assess the ability of each model to reliably predict independent and identically distributed data from $Y$, given all of the $X$ features versus all but one feature: $X_k$ (i.e. 'knock-out X'). This procedure results in two correlation metrics $R_{full}$ and $R_{knockout}$, whose difference $\Delta R=R_{full}-R_{knockout}$ indicates how much $X_i$ improves the prediction of $Y$ above and beyond what can be predicted from all other features $X_k$.

For, the Forward Model, this feature importance is assessed as follow:

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$,
          $X_{test} \in \mathbb{R}^{n' \times d_x}$,
          $Y_{train} \in \mathbb{R}^{n\times d_y}$,
          $Y_{test} \in \mathbb{R}^{n'\times d_y}$,
          }
    \KwOut{estimate of prediction improvement $\Delta{R} \in \mathbb{D}^{d_x, d_y}$.}
    $H = \text{LinearRegression}(X_{train}, Y_{train})$\;
    $R_{full} = \text{corr}(X_{test} H, Y_{test})$\;

    \For{$i = 1, \ldots, d_x$}{
        $K = Id$\;
        $K[i] \leftarrow 0$\;
        $H_k = \text{LinearRegression}(X_{train} K, Y_{train})$\;
        $R_{k} = \text{corr}(X_{test} {H_k}, Y_{test})$\;
        $\Delta R_i = R_{full} - R_{k}$\;
    }
    \Return{$\Delta R$}
    \caption{Forward feature importance.}
    \label{algorithm:fwd_fi}
\end{algorithm}

\iffalse
\begin{enumerate}
  \item Fit $H$ given $X_{train}$ and $Y_{train}$
  \item For each feature $i$:
  \begin{itemize}
    \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
    \item Fit $H^i_k$ given $X_{train} K$ and $Y_{train}$
  \end{itemize}
\end{enumerate}
\fi

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$,
          $X_{test} \in \mathbb{R}^{n' \times d_x}$,
          $Y_{train} \in \mathbb{R}^{n\times d_y}$,
          $Y_{test} \in \mathbb{R}^{n'\times d_y}$,
          }
    \KwOut{estimate of prediction improvement $\Delta{R} \in \mathbb{D}^{d_x, d_z}$.}
    $H, G = \text{CCA}(X_{train}, Y_{train})$\;
    $R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;

    \For{$i = 1, \ldots, d_x$}{
        $K = Id$\;
        $K[i] \leftarrow 0$\;
        $H_k, G_k = \text{CCA}(X_{train} K, Y_{train})$\;
        $R_{k} = \text{corr}(X_{test} H_k, Y_{test} G_k)$\;
        $\Delta R_i = R_{full} - R_{k}$\;
    }
    \Return{$\Delta R$}
    \caption{CCA and PLS feature importance.}
    \label{algorithm:cdp_fi}
\end{algorithm}

\iffalse
For CCA and PLS, feature importance is assessed as follow:
\begin{enumerate}
  \item Fit $H$ and $G$ given $X_{train}$ and $Y_{train}$
  \item For each feature $i$:
  \begin{itemize}
    \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
    \item Fit $H^i_k$ and $G_k$ given $X_{train} K$ and $Y_{train}$
  \end{itemize}
  \end{enumerate}
\fi

For the Backward Model, feature importance cannot be assessed. We thus simply report the model performance $R_i = corr(YG_i, X_i)$

For, B2B feature importance is assessed as follow:

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$,
          $X_{test} \in \mathbb{R}^{n' \times d_x}$,
          $Y_{train} \in \mathbb{R}^{n\times d_y}$,
          $Y_{test} \in \mathbb{R}^{n'\times d_y}$,
          }
    \KwOut{estimate of prediction improvement $\Delta{R} \in \mathbb{D}^{d_x}$.}
    $H, G = \text{B2B}(X_{train}, Y_{train})$\;
    $R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;

    \For{$i = 1, \ldots, d_x$}{
        $K = Id$\;
        $K[i] \leftarrow 0$\;
        $H_k = \text{LinearRegression}(X_{train} K, Y_{train} G_i)$\;
        $R_{k} = \text{corr}(X_{test} H_k, Y_{test} G_i)$\;
        $\Delta R_i = R_{full} - R_{k}$\;
    }
    \Return{$\Delta R$}
    \caption{B2B feature importance.}
    \label{algorithm:cdp_fi}
\end{algorithm}

\iffalse
\begin{enumerate}

  \item Fit $G$ given $X_{train}$ and $Y_{train}$
  \item Fit $H$ given $X_{train}$ and $Y_{train} G^T$
  \item For each feature $i$:
  \begin{itemize}
    \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
    \item Fit $H^i_k$ given $X_{train} K$ and $Y_{train} G^T$
  \end{itemize}

  \item Compute $R_{full}$ given $Y_{test} G^T$ and $X_{test}$
  \item For all $i \in d_x$:
  \begin{itemize}
     \item Compute $R^i_{knockout}$ given $Y_{test} G^T$ and $X_{test} K {H^i_k}^T$
     \item Compute $R^i_{delta}=R_{full} - R^i_{knockout}$
   \end{itemize}
\end{enumerate}
\fi

\iffalse
Ridge regression \citep{hoerl1959optimum}, multitask Lasso \citep{argyriou2008convex}, Partial Least Squares or PLS \citep{wold_pls, tenenhaus_pls}, Canonical Correlation Analysis or CCA \citep{cca_hotelling}, and Reduced Rank Regression or RRR \citep{Izenman_rrr}.
%
Each of these methods estimates a matrix of coefficients $\hat{W} \in \mathbb{R}^{d_x \times d_y}$, from which we estimate the diagonal mask $\hat{E}$ using the feature importances $\hat{E}_{i,i} = \| W_{i, :} \|$.
%
For B2B, we obtain $\hat{E}$ as described in Algorithm~\ref{algorithm:b2br}.
%

@@

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\textwidth]{synthetic.pdf}
    \caption{Results of synthetic experiments. Each dot depicts the average value of a metric for B2B ($x$-axis) against a competitor ($y$-axis) for each experiment configuration averaged over $20$ random seeds. The prediction error on $Y$ within (in) and outside (out) of the training distribution are weakly but consistently larger for all of the models as compared to B2B with and without the Sonquist binarization (E). The AUC, measuring the ability of each model to separate causal from noncausal factors, is consistently larger in B2B as compared to the other models.}
    \label{table:synthetic}
\end{figure}
\fi
